Tree size, (microhabitat)?, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2019-08-15 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

---
title: "Tree size and hydraulic traits shape drought responses in a temperate broadleaf forest"
author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: [book.bib,  packages.bib]
biblio-style: apalike
csl: apa.csl
header-includes:  
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \usepackage{dcolumn} #for sidewaystable
---

@@

*Study site*
Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38°53'36.6"N, 78° 08'43.4"W) [@andersonteixeira_ctfs-forestgeo:_2015]. SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.  Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013]. Dominant species include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).

@@

To accurately understand climate sensitivity, this study used a specific definition of drought, which is not a common practice [@slette_how_2019]. We used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined by [@lloret_components_2011] as the ratio between the performance during and before the disturbance. Candidate drought years were defined if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years. These were then cross-validated with the regional Palmer Drought Severity Index (PDSI) values for each year, which yielded a set of three periods that were consistently shown as drought: 1964-1966, 1977, and 1999.

### Results
Results outline

Descriptions of Droughts

- physical characteristics of drought: duration, min PDSI (and when occurred), see slette et al 2019 to see how they suggest reporting each one
- sum nb.series for each scenario, mult by perc.neg (pointers df), define thresholds (e.g. 75% of all trees experienced the growth reduction of 40% or more)
  - make table for this
  - citing table 3 (the coefficient for individual year) - maybe mention how 77 and 99 have neg coefficients, indicating potential height effect: as time goes on, trees are more affected by drought anyways because they grow taller
  - this is followed by Figure 2

Hypothesis results
- one header for each main question
- present results for each one, walk through via order and logic of Table 1
- no citations here

Once the data was collected, linear mixed models were run following the order of the hypotheses as seen in Figure ??? [individual_tested_traits]. Using the [@R-pointRes] package, we set up models with the resistance value as the response variable, and each prediction's variable as the independent variable. Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].Null models were determined in order of the predictions. First, we analyzed the combined scenario to determine if "year" was significant. Upon establishing this, we tested height and DBH as size parameters. Although both were significant, height was kept due to its larger delta AICc compared with the null model. We then tested the remaining biophysical and hydraulic traits individually against a null model containing height and year. This yielded Figure ??? (cand_full). All variables with dAICc >2 were used as candidates for each scenario's best model (figure ???? (tested_traits_best))

**Describe each individual drought scenario **
1964-1966
1977
1999

>>>>

---
title: "Tree size, (microhabitat)?, and hydraulic traits shape drought responses in a temperate broadleaf forest"
author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norm Bourg?, Bill McShea?, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: [book.bib,  packages.bib]
biblio-style: apalike
csl: apa.csl
header-includes:  
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \usepackage{dcolumn} #for sidewaystable
---

@@

*Study site*
Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38°53'36.6"N, 78° 08'43.4"W) [[Bourg et al. 2013](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/13-0010.1); @andersonteixeira_ctfs-forestgeo:_2015]. SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.  Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013]. Dominant species include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).

@@

We identified droughts within the time period as 1950-2009, defining drought as events where where peak growing season climatic conditions were anomalously dry relative to the long-term record and where tree growth was substantially reduced for at least some components of the community. To identify years with widespread reductions in tree growth, we used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined as the ratio between the performance during and before the disturbance [@lloret_components_2011]. Candidate drought years were defined if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years for at least three species-canopy class combinations, where canopy class included canopy (dominant and co-domiant) and subcanopy (suppressed and intermediate). **[this includes 1991- why did we drop it? [issue #52](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/52)]** Candidate droughts were then cross-validated with the regional Palmer Drought Severity Index (PDSI) values [**source?**] **(and MJJA PET-PRE (droughts identified in Helcoski)?)**. **[This paragraph needs work. See issue #52](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/52). Put climate criteria first.]**

*Analysis* 
Once the data was collected, linear mixed models were run following the order of the hypotheses as seen in Figure ??? [individual_tested_traits]. Using the [@R-pointRes] package, we set up models with the resistance value as the response variable, and each prediction's variable as the independent variable. Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].Null models were determined in order of the predictions. First, we analyzed the combined scenario to determine if "year" was significant. Upon establishing this, we tested height and DBH as size parameters. Although both were significant, height was kept due to its larger delta AICc compared with the null model. We then tested the remaining biophysical and hydraulic traits individually against a null model containing height and year. This yielded Figure ??? (cand_full). All variables with dAICc >2 were used as candidates for each scenario's best model (figure ???? (tested_traits_best))

### Results

*Descriptions of Droughts*
In the 60-year period between 1950 and 2009, there were three droughts that met our definition of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1964-1966, 1977, and 1999 (Fig. 1). We note that 1964 and 1966 were separately identified as drought years and grouped into a single multi-year drought period because 1965 was also anomalously hot and dry, albiet not identified as a pointer year. Climatically, these droughts included four of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019]. Specifically, 1964, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.87, 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively. The year 2007 also ranked in the top five driest (PET-PRE = 82.13 mm mo<sup>-1</sup>), but it did not have particualarly low PDSI and was not identified as a pointer year. **add stats for PDSI, T, precip**

- physical characteristics of drought: duration, min PDSI (and when occurred), see [@slette_how_2019] to see how they suggest reporting each one
- sum nb.series for each scenario, mult by perc.neg (pointers df), define thresholds (e.g. 75% of all trees experienced the growth reduction of 40% or more)

  - citing table 3 (the coefficient for individual year) - maybe mention how 77 and 99 have neg coefficients, indicating potential height effect: as time goes on, trees are more affected by drought anyways because they grow taller
